Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-09-28 Dail_Mad_ext.tex
intro

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$,
coupled with a stochastic model allowing
for random process variation. %, which we denote $\grave{o}_t$.
The deterministic model may be density-dependent, for example
logistic, or it might be density-independent, as in the case of
exponential growth.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcom of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are the most widespread approach for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects, (2) the model for observation error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative and non-integer values of the state-variable, which is
inconsistent with the observed data, and may result in implausible predictions.
%The fact that both abundance ($N_t$) and the observed data ($X_t$) are
%typically non-negative integers suggests that it is inappropriate to
%model process variation and observation error using Gaussian
%distributions.
Transforming the count data and the underlying abundance variables do
not fully resolve these issues \citep{ohara_kotze:2010}.
%Nonetheless, efforts have been made to adhere to the
%Gaussian assumptions by tranforming abundance to
%density by replacing $N_t$ with $Y_t = \log(N_t/A)$ where $A$ is the
%area surveyed. However, this transformation is problematic since
%local abundance may be zero and thus $Y_t = \log(0) = -\infty$. Generally,
%zeros are replaced with some small number, although the
%effect of this is rarely discussed.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation: \hl{[I'm not familiar with bracket notation,
  so these equations are confusing to me.  How widely used is this
  notation in ecology?  Will others likely understand this?  Also, we might want
to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. The three lines of general model
described by Eq.~\ref{eq:ss2} correspond to the three equations shown
for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
we allow $N_1$ to be a random variable. But what should its
distribution, $[N_1|\bm{\theta}]$ be? And how could $\bm{\theta}$ be
estimated since there is only a single observation available?
One approach to estimating the $\bm{\theta}$ is to assume that the
population is in equlibrium, i.e. that the expected value of $N_t$ is
constant through time. Althought this makes the parameters
identifiable, assuming equlibrium defeats the objective of many
studies of population dynamics, namely determining why a population
varies over time.

\begin{comment}
  However, a completSpecifically, the Markovian nature of the model
  implies the following likelihood The first term is the probability
  density for the first observation in the time series, $X_1$, yet it
  is intuitively not possible to estimate the parameter(s) of this
  distribution using only a single observation.
  % This problem has resulted in several papers with interesting names
  % like ``multi-modal likelihoods and ridges etc...''
  Two options have been proposed handling the identifiability
  problem. First, one can assume that the first observation in the
  time series is made without error. Second, one can assume that the
  population is at equilibrium so that that the x can be replaced with
  the equilibrium distribution. However, assuming equilibrium defeats
  the objective of many studies of population dynamics, namely
  determining if and why a population is at equilibrium.
\end{comment}

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
dependence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes. In contrast, recent work has sought to use state-space
models for inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.

@@

>>>>

A simple state-space models can be described as follows.
Let $N_t$ be the abundance of a species during year $t$, for
$t=1,\hdots,T$, and let $X_t$ be
the observed data, which differs from $N_t$ due to observation error,
a random effect denoted $\zeta_t$. Temporal variation in $N_t$ is
modeled using a population growth model, $\mu(N_{t-1})$
and random process variation denoted $\grave{o}_t$.
The growth model may be density-dependent, as in the case of the
Ricker model, or it might be density-independent, such as when growth
is exponential.
%, in which $\mu(N_{t-1}) = N_{t-1}e^{r}$ where $r$ is the intrinsic rate of increase.
The full model can now be written as:
\begin{subequations}
  \label{eq:ss1}
  \begin{align}
    N_1 &= X_1 \label{eq:ss1a} \\
    N_t &= \mu(N_{t-1}) + \grave{o}_{t-1} \quad \text{for} \; t=2,\hdots,T \label{eq:ss1b} \\
    X_t &= N_t + \zeta_t \qquad \qquad \;\, \text{for} \; t=1,\hdots,T \label{eq:ss1c}
  \end{align}
\end{subequations}
where $\grave{o}_t$ is the random effect allowing for process
variation unaccounted for by the deterministic model. In classical
state-space models, the two sets of random effects
are assumed to be i.i.d. Gaussian deviates:
$\grave{o}_t \sim \mathrm{N}(0, \sigma)$ and
$\zeta_t \sim \mathrm{N}(0, \tau)$. It is also standard practice to
ignore process variation associated with $N_1$, as
indicated by Eq~\ref{eq:ss1a}. Alternatively, the population is
assumed to be at equilibrium such that the $N_1$ can be regarded as an
outcom of the equilibrium distribution.

Even though state-space models such as that shown in Eq.~\ref{eq:ss1}
are among the most widespread approaches for modeling population dynamics
using time-series data, several problems are evident. Namely, (1)
time-series data and the underlying abundance parameters of interest
are typically integer-valued, as in the case of count data, raising
concerns about the use of Gaussian distribution for the random
effects, (2) the model for observation error has little biological basis, (3)
spatial variation in abundance is not allowed, and (4) some of the
paramters of the model are not estimable. We briefly discuss each of
these points before describing a general approach to resolving these
shortcomings.

Use of the Gaussian distribution for modeling random process variation
and observation error is motivated by convenience rather than
biology. Specifically, the Gaussian assumptions
allow for parameter estimation using the Kalman filter
\citep{dennis_etal:2006}, which is much more computationally efficient
than estimation methods when random effects are not Gaussian distributed
\citep{devalpine_hastings:2002}. The problem with this is that it
allows for negative and non-integer values of the state-variable, which is
inconsistent with the observed data, and may result in implausible predictions.
Transforming the count data and the underlying abundance variables do
little to resolve these issues \citep{ohara_kotze:2010}.

@@

A more serious problem than the ones associated with the Gaussian
assumptions is that the parameters of the simple state-space models
such as Eq~\ref{eq:ss1} are not identifiable in many
circumstances \citep{polansky_etal:2009}. Recall that the
Eq.~\ref{eq:ss1} did not specify distributions for $N_1$; % or $X_1$;
however, $N_1$ is a random variable and hence there is uncertainty
that should be accounted for.
%are needed to fully specify the model.
To be more specific, a fully-specified state-space model requires at least
three probability distributions, which we represent
using bracket notation (i.e., $[Y|\Omega]$ is interpretted
as the probability distribution of the random variable $Y$ given the
parameter $\Omega$). The three probability distributions required for a
state-space model are:
%\hl{[I'm not familiar with bracket notation,
%  so these equations are confusing to me.  How widely used is this
%  notation in ecology?  Will others likely understand this?  Also, we might want
%to save space by trimming a bunch of this discussion back down.]}
\begin{subequations}
  \label{eq:ss2}
  \begin{align}
    [N_1&|\bm{\theta}] \\ \label{eq:ss2b}
    [N_t&|N_{t-1},\bm{\Theta}] \quad \text{for} \; t=2,\hdots,T \\ \label{eq:ss2c}
    [X_t&|N_t,\bm{p}]  \qquad \; \text{for} \; t=1,\hdots,T
  \end{align}
\end{subequations}
where $\bm{\theta}$ are the process variation parameters for the
initial state, $\bm{\Theta}$ are the process variation parameters
influencing how abundance changes over time, and $\bm{p}$ are the
observation error parameters. As mentioned previously, Eq.~\ref{eq:ss2b}
and Eq.~\ref{eq:ss2c} are assumed to be Gaussian in classical state-space
models, but what should be the distribution for Eq.~\ref{eq:ss2}?
%The three lines of general model
%described by Eq.~\ref{eq:ss2} correspond to the three equations shown
%for the specific example in Eq~\ref{eq:ss1}.  However, in this case,
%we allow $N_1$ to be a random variable. But what should its
%istribution, $[N_1|\bm{\theta}]$ be?
And how could $\bm{\theta}$ be
estimated given that there is only a single observation available?
Often, these problems are simply assumed away. For
example, sometimes researchers assume that there is no process variation in
the first year, although this seems unjustified. Alternatively, the
population may be assumed to be at equilibrium such that the expected
value of $N_t$ is constant through time. Although this
makes the parameters $\bm{\theta}$ identifiable, assuming equlibrium
defeats the objective of many studies of population dynamics, namely
determining why a population varies over time.

@@

Several extensions of state-space models have been proposed to
overcome the limitations described
above. \citet{devalpine_hastings:2002} and \citet{kery_etal:2009}
described methods for fitting models with non-Gaussian distributions
for the process and observation errors. Observation models with more
intuitive interpretations, such as those that explicitly model
detection probability, have been proposed by
\citet{kery_etal:2009}. \citet{lele_etal:1998} and
\citet{kery_etal:2009} developed models allowing for inference about
spatial and temporal variation in abundance, and their developments
also resolved the problems of non-identifiability for the parameters
of the initial state at time $t=1$. Of these extensions, the work by
\citet{kery_etal:2009} is unique in that it addressed multiple limitations
simultaneously. However, their model did not include serial
dependence, which is a hallmark of population models. This limits the
utility of their model for making inferences about explicit population
processes.

@@

In contrast, recent work has sought to use state-space
models for inference about population processes such as mortality and
recruitment. To do so, methods have been developed to combine multiple
sources of information, such as counts and mark-recapture data
\citep{besbeas_etal:2002, buckland_etal:2004,
  schaub_etal:2007}. Although these integrated
population models may be viewed as the gold standard in state-space
modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the world,
such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.
